\Large Taking uncertainty seriously \newline \normalsize A Bayesian approach to word embedding bias estimation

### 2022-03-21 presentations/presentationBoston/presentationBoston.tex
Adds final version

\begin{frame}{Presentation plan}
\protect\hypertarget{presentation-plan}{}
\begin{itemize}
\tightlist
\item
  Bias in word embeddings
\item
  WEAT and MAC methods
\item
  Methodological problems
\item
  Limitations of pre-averaging in bias detection methods
\item
  Accounting for uncertainty with Bayesian approach
\end{itemize}
\end{frame}

\begin{frame}{Word2vec}
\protect\hypertarget{word2vec}{}
\begin{block}{Question}
\protect\hypertarget{question}{}
How to sensibly represent words with numbers?

@@

\dots \dots but this would be inefficient and wouldn't capture any
relations between words.
\end{block}
\end{frame}

\begin{frame}{Word2vec}
\protect\hypertarget{word2vec-1}{}
\begin{center}
 \includegraphics[width = 9cm]{images/perceptron1.png}
\end{center}

@@

\begin{frame}{Word2vec}
\protect\hypertarget{word2vec-2}{}
\begin{center}
 \includegraphics[width = 9cm]{images/perceptron1.png}
\end{center}

@@

\begin{frame}{Word2vec}
\protect\hypertarget{word2vec-3}{}
\begin{columns}
\column{0.45\linewidth}

\begin{center}
 \includegraphics[height = 6cm, width = 5cm]{images/perceptron2.png}
\end{center}

\vspace{-3mm}
\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}

@@

\begin{frame}{Word2vec}
\protect\hypertarget{word2vec-4}{}
\begin{block}{Distributional semantics}
\protect\hypertarget{distributional-semantics}{}
\begin{itemize}
\item "You shall know a word by the company it keeps" (John Firth, 1957)

\item "the degree of semantic similarity between two linguistic expressions $A$ and $B$ is a function of the similarity of the linguistic contexts in which $A$ and $B$ can appear." (A. Lenci, 2008)

@@

\begin{frame}{Word2vec}
\protect\hypertarget{word2vec-5}{}
\begin{block}{Google and Mikolov}
\protect\hypertarget{google-and-mikolov}{}
\emph{Efficient Estimation of Word Representation in Vector Space}, 2013

@@

\begin{frame}{Word2vec}
\protect\hypertarget{word2vec-6}{}
\begin{block}{Google and Mikolov}
\protect\hypertarget{google-and-mikolov-1}{}
\emph{Efficient Estimation of Word Representation in Vector Space}, 2013

@@

\begin{frame}{Word2vec}
\protect\hypertarget{word2vec-7}{}
\begin{longtable}[]{@{}llllll@{}}
\toprule
word & 1 & 2 & 3 & 4 & \ldots{} \\
\midrule
\endhead
woman & 0.456 & 0.267 & 0.675 & 0.131 & \ldots{} \\
man & 0.451 & 0.897 & 0.472 & 0.088 & \ldots{} \\
\bottomrule
\end{longtable}

@@

\begin{center}\includegraphics[width=1\linewidth]{presentationBoston_files/figure-beamer/cosine1-1} \end{center}

@@

\begin{center}\includegraphics[width=1\linewidth]{presentationBoston_files/figure-beamer/cosine2-1} \end{center}

@@

\begin{center}\includegraphics[width=1\linewidth]{presentationBoston_files/figure-beamer/cosine3-1} \end{center}

@@

\begin{center}\includegraphics[width=1\linewidth]{presentationBoston_files/figure-beamer/cosine4-1} \end{center}

@@

  \begin{itemize}
  \tightlist
  \item
    \(1\): maximally smilar
  \item
    \(-1\): opposites
  \item
    \(0\): dissimilar
  \end{itemize}
\item
  \(\mathsf{cosineDistance}\in (0, 2)\)
\end{itemize}
\end{frame}

@@

\begin{block}{No design considerations to sample size}
\protect\hypertarget{no-design-considerations-to-sample-size}{}
\begin{itemize}
\tightlist
\item
  \protect\hyperlink{ref-Ethayarajh2020measuring}{Ethayarajh}
  (\protect\hyperlink{ref-Ethayarajh2020measuring}{2020}) criticizes
  WEAT uses Bernstein bounds and argues that we would need a bias
  specific dataset of size at least 11903 to claim that the system is
  biased (three times larger than WinoBias).
\end{itemize}

@@

\begin{frame}{Some methodological problems}
\protect\hypertarget{some-methodological-problems-1}{}
\begin{block}{No word class distinction and no control group}
\protect\hypertarget{no-word-class-distinction-and-no-control-group}{}
We make the subclasses clear, add human neutral predicates and neutral
predicates for control. We used L2-Reddit corpus and GoogleNews (we
present the results for Reddit for brevity).

@@

\begin{frame}{Some methodological problems}
\protect\hypertarget{some-methodological-problems-4}{}
\begin{block}{Distances for ``muslim''}
\protect\hypertarget{distances-for-muslim}{}
\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-4-1} \end{center}
\end{block}
\end{frame}

\begin{frame}{Some methodological problems}
\protect\hypertarget{some-methodological-problems-5}{}
\begin{block}{Distances for ``priest''}
\protect\hypertarget{distances-for-priest}{}
\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-5-1} \end{center}
\end{block}
\end{frame}

\begin{frame}{Some methodological problems}
\protect\hypertarget{some-methodological-problems-6}{}
\begin{block}{No principled interpretation}
\protect\hypertarget{no-principled-interpretation}{}
\begin{longtable}[]{@{}lllll@{}}
\toprule
Category & Biased & Hard Debiased & Diff & \\
\midrule
\endhead
Religion & 0.859 & 0.934 & 0,141 & \\
Race & 0.892 & 0.925 & 0,108 & \\
Gender & 0.623 & 0.700 & 0,377 & \\
\bottomrule
\end{longtable}

@@

\begin{frame}{Some methodological problems}
\protect\hypertarget{some-methodological-problems-7}{}
\begin{block}{No principled interpretation - visualization 1}
\protect\hypertarget{no-principled-interpretation---visualization-1}{}
\begin{center}\includegraphics[width=0.9\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-6-1} \end{center}
\end{block}
\end{frame}

\begin{frame}{Some methodological problems}
\protect\hypertarget{some-methodological-problems-8}{}
\begin{block}{No principled interpretation - visualization 2}
\protect\hypertarget{no-principled-interpretation---visualization-2}{}
\begin{center}\includegraphics[width=0.9\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-7-1} \end{center}
\end{block}
\end{frame}

\begin{frame}{Some methodological problems}
\protect\hypertarget{some-methodological-problems-9}{}
\begin{block}{No principled interpretation - visualization 3}
\protect\hypertarget{no-principled-interpretation---visualization-3}{}
\begin{center}\includegraphics[width=0.9\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-8-1} \end{center}
\end{block}
\end{frame}

@@

\begin{frame}{The problem with pre-averaging}
\protect\hypertarget{the-problem-with-pre-averaging-1}{}
\begin{block}{Simple case: two pws, four terms}
\protect\hypertarget{simple-case-two-pws-four-terms}{}
\vspace{1mm}
\footnotesize

\begin{center}\includegraphics[width=0.7\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-9-1} \end{center}
\normalsize

@@

\begin{itemize}
\tightlist
\item
  Raw sd in data is 0.072
\item
  The sd of means is 0.037
\item
  The WEAT score is 1.825\\
\item
  The largest effect size reported by
  \protect\hyperlink{ref-Caliskan2017semanticsBiases}{Caliskan, Bryson,
  \& Narayanan}
  (\protect\hyperlink{ref-Caliskan2017semanticsBiases}{2017}) is 1.81!
\end{itemize}
\end{block}
\end{frame}

\begin{frame}{The problem with pre-averaging on realistic set-up}
\protect\hypertarget{the-problem-with-pre-averaging-on-realistic-set-up}{}
\begin{block}{Simulation with realistic set-up (16 predicates)}
\protect\hypertarget{simulation-with-realistic-set-up-16-predicates}{}
\vspace{1mm}
\footnotesize

\begin{center}\includegraphics[width=0.7\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-11-1} \end{center}

@@

Raw sd: 0.082, sd of means: 0.031, WEAT: 2.337.
\end{block}
\end{frame}

\begin{frame}{The problem with pre-averaging on realistic set-up}
\protect\hypertarget{the-problem-with-pre-averaging-on-realistic-set-up-1}{}
\begin{block}{10k simulations (same parameters)}
\protect\hypertarget{k-simulations-same-parameters}{}
\vspace{1mm}
\footnotesize

\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-12-1} \end{center}
\vspace{1mm}
\footnotesize

@@

\begin{itemize}
\tightlist
\item
  95\% of the scores are in range -2.763, 2.698
\item
  21.38\% of the absolute values are above 1.81
\end{itemize}
\end{block}
\end{frame}

\begin{frame}{The problem with pre-averaging on realistic set-up}
\protect\hypertarget{the-problem-with-pre-averaging-on-realistic-set-up-2}{}
\begin{block}{10k simulations with mean similarity 0.1}
\protect\hypertarget{k-simulations-with-mean-similarity-0.1}{}
\vspace{1mm}
\footnotesize

\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-14-1} \end{center}
\vspace{1mm}
\footnotesize

@@

\begin{itemize}
\tightlist
\item
  95\% of the scores are in range 0.851, 2.764
\item
  51.3\% of the absolute values are above 1.81
\end{itemize}
\end{block}
\end{frame}

\begin{frame}{The problem with pre-averaging on realistic set-up}
\protect\hypertarget{the-problem-with-pre-averaging-on-realistic-set-up-3}{}
\begin{block}{10k simulations with mean similarity 0.4}
\protect\hypertarget{k-simulations-with-mean-similarity-0.4}{}
\vspace{1mm}
\footnotesize

\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-16-1} \end{center}
\vspace{1mm}
\footnotesize

@@

\begin{itemize}
\tightlist
\item
  95\% of the scores are in range 1.679, 2.185
\item
  82.9\% of the absolute values are above 1.81
\end{itemize}
\end{block}
\end{frame}

@@

\begin{frame}{Thank you!}
\protect\hypertarget{thank-you}{}
\begin{block}{Summary}
\protect\hypertarget{summary}{}
\begin{itemize}
\tightlist
\item
  Bias in word embeddings
\item
  WEAT and MAC methods
\item
  Methodological problems
\item
  Limitations of pre-averaging in bias detection methods
\item
  Accounting for uncertainty with Bayesian approach
\end{itemize}
\end{block}

\begin{block}{Further work}
\protect\hypertarget{further-work}{}
\begin{itemize}
\tightlist
\item
  Including contrasts in Bayesian calculation
\item
  Performance cross-validation in comparison to other methods (regular
  linear regression, KNN, \dots)
\item
  Downstream tasks and connection with intrinsic evaluation
\item
  Testing data from the original Implicit Association Test (IAT)
\item
  Applying uncertainty to WEAT and better word lists
\item
  Looking at other similarity measures
\end{itemize}
\end{block}
\end{frame}

>>>>

\begin{frame}{Presentation plan}
\protect\hypertarget{presentation-plan}{}
\begin{itemize}
\tightlist
\item
  Word embeddings
\item
  Cosine similarity
\item
  Bias in word embeddings
\item
  WEAT and MAC measures
\item
  Methodological problems
\item
  Bad consequences of pre-averaging
\item
  Our Bayesian alternative
\end{itemize}
\end{frame}

\begin{frame}{Word embeddings}
\protect\hypertarget{word-embeddings}{}
\begin{block}{Question}
\protect\hypertarget{question}{}
How to sensibly represent words with numbers?

@@

\dots but this would be inefficient and wouldn't capture any relations
between words.
\end{block}
\end{frame}

\begin{frame}{Word embeddings}
\protect\hypertarget{word-embeddings-1}{}
\begin{center}
 \includegraphics[width = 9cm]{images/perceptron1.png}
\end{center}

@@

\begin{frame}{Word embeddings}
\protect\hypertarget{word-embeddings-2}{}
\begin{center}
 \includegraphics[width = 9cm]{images/perceptron1.png}
\end{center}

@@

\begin{frame}{Word embeddings}
\protect\hypertarget{word-embeddings-3}{}
\begin{columns}
\column{0.45\linewidth}

\begin{center}
 \includegraphics[height = 5.5cm, width = 5cm]{images/perceptron2.png}
\end{center}

\vspace{2mm}
\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}

@@

\begin{frame}{Word embeddings}
\protect\hypertarget{word-embeddings-4}{}
\begin{block}{Distributional semantics}
\protect\hypertarget{distributional-semantics}{}
\begin{itemize}
\item "You shall know a word by the company it keeps" (John Firth, 1957)

\item "\dots the degree of semantic similarity between two linguistic expressions $A$ and $B$ is a function of the similarity of the linguistic contexts in which $A$ and $B$ can appear." (A. Lenci, 2008)

@@

\begin{frame}{Word embeddings}
\protect\hypertarget{word-embeddings-5}{}
\begin{block}{Google and Mikolov}
\protect\hypertarget{google-and-mikolov}{}
\emph{Efficient Estimation of Word Representation in Vector Space}, 2013

@@

\begin{frame}{Word embeddings}
\protect\hypertarget{word-embeddings-6}{}
\begin{block}{Google and Mikolov}
\protect\hypertarget{google-and-mikolov-1}{}
\emph{Efficient Estimation of Word Representation in Vector Space}, 2013

@@

\begin{frame}{Word embeddings}
\protect\hypertarget{word-embeddings-7}{}
\begin{longtable}[]{@{}llllll@{}}
\toprule
word & 1 & 2 & 3 & 4 & \ldots{} \\
\midrule
\endhead
woman & 0.456 & 0.267 & 0.675 & 0.131 & \ldots{} \\
man & 0.451 & 0.897 & 0.472 & 0.088 & \ldots{} \\
\bottomrule
\end{longtable}

@@

\begin{center}\includegraphics[width=1.15\linewidth]{presentationBoston_files/figure-beamer/cosine1-1} \end{center}

@@

\begin{center}\includegraphics[width=1.15\linewidth]{presentationBoston_files/figure-beamer/cosine2-1} \end{center}

@@

\begin{center}\includegraphics[width=1.15\linewidth]{presentationBoston_files/figure-beamer/cosine3-1} \end{center}

@@

\begin{center}\includegraphics[width=1.15\linewidth]{presentationBoston_files/figure-beamer/cosine4-1} \end{center}

@@

  \begin{itemize}
  \tightlist
  \item
    \(1\): maximally smilar
  \item
    \(-1\): opposites
  \item
    \(0\): lack of similarity
  \end{itemize}
\item
  \(\mathsf{cosineDistance}\in (0, 2)\)
\end{itemize}
\end{frame}

@@

\begin{block}{No design considerations to sample size}
\protect\hypertarget{no-design-considerations-to-sample-size}{}
\begin{itemize}
\tightlist
\item
  \protect\hyperlink{ref-Ethayarajh2020measuring}{Ethayarajh}
  (\protect\hyperlink{ref-Ethayarajh2020measuring}{2020}) uses Bernstein
  bounds to criticize WEAT, and argues that we would need a bias
  specific dataset of size at least 11903 to claim that the system is
  biased (three times larger than WinoBias).
\end{itemize}

@@

\begin{frame}{Some methodological problems}
\protect\hypertarget{some-methodological-problems-1}{}
\begin{block}{No word class distinction and no control group}
\protect\hypertarget{no-word-class-distinction-and-no-control-group}{}
We make the subclasses clear, add human neutral predicates and neutral
predicates for control. We used L2-Reddit corpus and GoogleNews
\linebreak \footnotesize (we present the results for Reddit for
brevity). \normalsize

@@

\begin{frame}{Some methodological problems}
\protect\hypertarget{some-methodological-problems-4}{}
\begin{center}\includegraphics[width=1\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-4-1} \end{center}
\end{frame}

\begin{frame}{Some methodological problems}
\protect\hypertarget{some-methodological-problems-5}{}
\begin{center}\includegraphics[width=1\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-5-1} \end{center}
\end{frame}

\begin{frame}{Some methodological problems}
\protect\hypertarget{some-methodological-problems-6}{}
\begin{block}{No principled interpretation}
\protect\hypertarget{no-principled-interpretation}{}
\begin{longtable}[]{@{}lllll@{}}
\toprule
Category & Biased & Hard Debiased & Diff & \\
\midrule
\endhead
Religion & 0.859 & 0.934 & 0.141 & \\
Race & 0.892 & 0.925 & 0.108 & \\
Gender & 0.623 & 0.700 & 0.377 & \\
\bottomrule
\end{longtable}

@@

\begin{frame}{Some methodological problems}
\protect\hypertarget{some-methodological-problems-7}{}
\begin{block}{No principled interpretation}
\protect\hypertarget{no-principled-interpretation-1}{}
\begin{center}\includegraphics[width=0.9\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-6-1} \end{center}
\end{block}
\end{frame}

\begin{frame}{Some methodological problems}
\protect\hypertarget{some-methodological-problems-8}{}
\begin{block}{No principled interpretation}
\protect\hypertarget{no-principled-interpretation-2}{}
\begin{center}\includegraphics[width=0.9\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-7-1} \end{center}
\end{block}
\end{frame}

\begin{frame}{Some methodological problems}
\protect\hypertarget{some-methodological-problems-9}{}
\begin{block}{No principled interpretation}
\protect\hypertarget{no-principled-interpretation-3}{}
\begin{center}\includegraphics[width=0.9\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-8-1} \end{center}
\end{block}
\end{frame}

@@

\begin{frame}{The problem with pre-averaging}
\protect\hypertarget{the-problem-with-pre-averaging-1}{}
\vspace{1mm}
\footnotesize

\begin{center}\includegraphics[width=0.95\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-9-1} \end{center}
\normalsize

@@

\begin{itemize}
\tightlist
\item
  Raw sd in data is 0.072, the sd of means is 0.037
\item
  The WEAT score is 1.825, the largest effect size reported by
  \protect\hyperlink{ref-Caliskan2017semanticsBiases}{Caliskan, Bryson,
  \& Narayanan}
  (\protect\hyperlink{ref-Caliskan2017semanticsBiases}{2017}) is 1.81!
\end{itemize}
\end{frame}

\begin{frame}{The problem with pre-averaging on realistic set-up}
\protect\hypertarget{the-problem-with-pre-averaging-on-realistic-set-up}{}
\vspace{1mm}
\footnotesize

\begin{center}\includegraphics[width=0.95\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-11-1} \end{center}

@@

Raw sd: 0.082, sd of means: 0.031, WEAT: 2.337.
\end{frame}

\begin{frame}{The problem with pre-averaging on realistic set-up}
\protect\hypertarget{the-problem-with-pre-averaging-on-realistic-set-up-1}{}
\vspace{1mm}
\footnotesize

\begin{center}\includegraphics[width=0.9\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-12-1} \end{center}
\vspace{1mm}
\footnotesize

@@

\begin{itemize}
\tightlist
\item
  95\% of the scores are in range -2.763, 2.698
\item
  21.38\% of the absolute values are above 1.81
\end{itemize}
\end{frame}

\begin{frame}{The problem with pre-averaging on realistic set-up}
\protect\hypertarget{the-problem-with-pre-averaging-on-realistic-set-up-2}{}
\vspace{1mm}
\footnotesize

\begin{center}\includegraphics[width=0.9\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-14-1} \end{center}
\vspace{1mm}
\footnotesize

@@

\begin{itemize}
\tightlist
\item
  95\% of the scores are in range 0.851, 2.764
\item
  51.3\% of the absolute values are above 1.81
\end{itemize}
\end{frame}

\begin{frame}{The problem with pre-averaging on realistic set-up}
\protect\hypertarget{the-problem-with-pre-averaging-on-realistic-set-up-3}{}
\vspace{1mm}
\footnotesize

\begin{center}\includegraphics[width=0.9\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-16-1} \end{center}
\vspace{1mm}
\footnotesize

@@

\begin{itemize}
\tightlist
\item
  95\% of the scores are in range 1.679, 2.185
\item
  82.9\% of the absolute values are above 1.81
\end{itemize}
\end{frame}

@@

\begin{frame}{Thank you!}
\protect\hypertarget{thank-you}{}
\begin{block}{Summary}
\protect\hypertarget{summary}{}
\pause

\begin{itemize}
\tightlist
\item
  The existing measures of bias by pre-averaging oversimplify the
  picture and lead to false confidence
\end{itemize}

\begin{itemize}
\tightlist
\item
  Our simulation results show that the mainstream interpretations are
  too hasty
\end{itemize}

\begin{itemize}
\item
  Bayesian analysis with proper control groups suggests that:

  \begin{itemize}
  \tightlist
  \item
    The word lists in use are too short
  \item
    Associated and different groups behave similarly, which undermines
    the semantic interpretation of cosine similarity
  \item
    Analysis at different levels of generality shows that there is more
    variety than one may initially think
  \end{itemize}
\end{itemize}

\begin{block}{Please remember about the feedback!}
\protect\hypertarget{please-remember-about-the-feedback}{}
\end{block}
\end{frame}
